---
title: "Identifying Selected Genes"
author: "Jennifer Blanc"
date: "2/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(reshape2)
library(ggpubr)
```

## Intro 

Here is the code used to identify selected genes in each of the 9 tissue types and make Figure 2. 

## Code 

The code below will be written to identify selected genes in the Kernel tissue. This code can be re-run for other tissues by specifying the path to the expression for that tissue. 

```{r}
# Read in mean-centered expression values
df1 <- read.table("../data/Mean_centered_expression/Kern.txt")
# Read in tissue specific kinship matrix 
myF <- read.table('../data/Kinship_matrices/F_Kern.txt')
```

First, we will calculate the Eigen values and vectors.
```{r}
## Get Eigen Values and Vectors 
myE <- eigen(myF)
E_vectors <- myE$vectors
E_values <- myE$values
```

Next, we will multiply the mean centered expression value by each of the m eigen vectors (numerator of equation 2). This step takes a little time (~10 minutes) so I will pre-load the results to speed up the knitting process 
```{r}
#df2 <- data.frame(matrix(ncol=ncol(df1), nrow=nrow(df1)))
#colnames(df2) <- colnames(df1[1:ncol(df1)])
#rownames(df2) <- rownames(df1)

## Calculate Q values by multiplying the mean-centered expression value by each eigen vector 
#for (i in 1:ncol(df2)) {
#  mean_centered_data <- t(as.matrix(as.numeric(df1[,i])))
#  for (k in 1:nrow(df2)){
#    u <- as.matrix(as.numeric(E_vectors[,k]))
#    value <- mean_centered_data %*% u
#    df2[k,i] <- value
#  }
#}
df2 <- read.table("../output/Identifying_Selected_Genes/intermediate_df2.txt")
```

Now divide by the square root of the eigenvalue to get the $C_m$ value from equation 2.  
```{r}
## Get the square root of the Eigen values   
de <- data.frame(matrix(nrow = nrow(df1),ncol = 2))
de$Egien_values <- E_values 
de$Sqrt_EV <- sqrt(de$Egien_values)

## Calculate C-values by dividing Q values by the square root of the eigen values
df4 <- data.frame(matrix(ncol=ncol(df2),nrow=nrow(df2)))
for (i in 1:ncol(df2)){
  df4[,i] <- (df2[,i] / de$Sqrt_EV)
}
```


We want to get the F value (equation 3) and the corresponding p-value for each of the first 5 PC's individually 
```{r}
## Calculate F-values by dividing variances 
F_values <- data.frame(matrix(ncol=ncol(df2), nrow = 1))
for (j in 1:ncol(df2)){
  for (i in 1:5){
    q <- df4[i,j] 
    t <- df4[11:20,j] 
    var_q <- mean(q^2)
    var_t <- mean(t^2)
    F_value <- var_q / var_t
    F_values[i,j] <- F_value
  }
}

## Calculate P-values from recorded F values 
P_values_ind <- data.frame(matrix(ncol=ncol(df2), nrow =1))
for (j in 1:ncol(F_values)){
  for (r in 1:5) {
    f_stat <- F_values[r, j]
    p_value <- pf(q=f_stat, df1=1, df2=10, lower.tail=FALSE) 
    P_values_ind[r, j] <- p_value
  }
}
```

Finally, we want to get the F value for the first 5 PC's combined (equation 4)  
```{r}
## Calculate F-values by dividing variances 
F_values <- data.frame(matrix(ncol=ncol(df2), nrow = 1))
for (j in 1:ncol(df2)){
  for (i in 1:1){
    q <- df4[1:5,j] 
    t <- df4[11:20,j]
    var_q <- mean(q^2)
    var_t <- mean(t^2)
    F_value <- var_q / var_t
    F_values[i,j] <- F_value
  }
}

## Calculate P-values from recorded F values 
P_values_comb <- data.frame(matrix(ncol=ncol(df2), nrow =1))
for (j in 1:ncol(F_values)){
  for (r in 1:1) {
    f_stat <- F_values[r, j]
    p_value <- pf(q=f_stat, df1=5, df2=10, lower.tail=FALSE) 
    P_values_comb[r, j] <- p_value
  }
}
```

The code above can be used to generate the p-values for all 9 tissues, the running this code is in "output/Identifying_Selected_Genes/". The files labeled "ALL" have the p-values for the first 5 PC's individually. The files labeled "Combined" have the p-values for the first 5 PC's combined.  

## Figure  

Read in the P-values for the first 5 PCs for each tissue 
```{r}
kern <- t(read.table("../output/Identifying_Selected_Genes/P_values_ALL_Kern.txt"))
gshoot <- t(read.table("../output/Identifying_Selected_Genes/P_values_ALL_GShoot.txt"))
groot <- t(read.table("../output/Identifying_Selected_Genes/P_values_ALL_GRoot.txt"))
base <- t(read.table("../output/Identifying_Selected_Genes/P_values_ALL_L3Base.txt"))
tip <- t(read.table("../output/Identifying_Selected_Genes/P_values_ALL_L3Tip.txt"))
lmad8 <- t(read.table("../output/Identifying_Selected_Genes/P_values_ALL_LMAD8.txt"))
lman8 <- t(read.table("../output/Identifying_Selected_Genes/P_values_ALL_LMAN8.txt"))
lmad26 <- t(read.table("../output/Identifying_Selected_Genes/P_values_ALL_LMAD26.txt"))
lman26 <- t(read.table("../output/Identifying_Selected_Genes/P_values_ALL_LMAN26.txt"))
```

Calculate the Q-values for each PC in each tissue and record the number of genes with Q-value < 0.1
```{r}
# Create table to collect selected genes
FDR <- as.data.frame(matrix(ncol=6, nrow=9))
colnames(FDR) <- c("PC1", "PC2", "PC3", "PC4", "PC5", "PC1:5")
rownames(FDR) <- c("Kern", "GShoot", "GRoot", "L3Base", "L3Tip", "LMAD8", "LMAN8", "LMAD26", "LMAN26")

# For each tissue calculate q-values and record the number of genes with q-values (and raw p-values) under 0.1
for (i in 1:5) {
  qobj <- qvalue::qvalue(kern[,i])
  FDR[1,i] <- length(qobj$qvalues[qobj$qvalues < 0.1])
}

for (i in 1:5) {
  qobj <- qvalue::qvalue(gshoot[,i])
  FDR[2,i] <- length(qobj$qvalues[qobj$qvalues < 0.1])
}

for (i in 1:5) {
  qobj <- qvalue::qvalue(groot[,i])
  FDR[3,i] <- length(qobj$qvalues[qobj$qvalues < 0.1])
}

for (i in 1:5) {
  qobj <- qvalue::qvalue(base[,i])
  FDR[4,i] <- length(qobj$qvalues[qobj$qvalues < 0.1])
}

for (i in 1:5) {
  qobj <- qvalue::qvalue(tip[,i])
  FDR[5,i] <- length(qobj$qvalues[qobj$qvalues < 0.1])
}

for (i in 1:5) {
  qobj <- qvalue::qvalue(lmad8[,i])
  FDR[6,i] <- length(qobj$qvalues[qobj$qvalues < 0.1])
}

for (i in 1:5) {
  qobj <- qvalue::qvalue(lman8[,i])
  FDR[7,i] <- length(qobj$qvalues[qobj$qvalues < 0.1])
}

for (i in 1:5) {
  qobj <- qvalue::qvalue(lmad26[,i])
  FDR[8,i] <- length(qobj$qvalues[qobj$qvalues < 0.1])
}

for (i in 1:5) {
  qobj <- qvalue::qvalue(lman26[,i])
  FDR[9,i] <- length(qobj$qvalues[qobj$qvalues < 0.1])
}
```

Read in P-values for PC1:5 combined for each tissue
```{r}
kern2 <- t(read.table("../output/Identifying_Selected_Genes/P_values_Combined_Kern.txt"))
gshoot2 <- t(read.table("../output/Identifying_Selected_Genes/P_values_Combined_GShoot.txt"))
groot2 <- t(read.table("../output/Identifying_Selected_Genes/P_values_Combined_GRoot.txt"))
base2 <- t(read.table("../output/Identifying_Selected_Genes/P_values_Combined_L3Base.txt"))
tip2 <- t(read.table("../output/Identifying_Selected_Genes/P_values_Combined_L3Tip.txt"))
lmad82 <- t(read.table("../output/Identifying_Selected_Genes/P_values_Combined_LMAD8.txt"))
lman82 <- t(read.table("../output/Identifying_Selected_Genes/P_values_Combined_LMAN8.txt"))
lmad262 <- t(read.table("../output/Identifying_Selected_Genes/P_values_Combined_LMAD26.txt"))
lman262 <- t(read.table("../output/Identifying_Selected_Genes/P_values_ALL_LMAN26.txt"))
```

Record number of genes with qvalues < 0.1
```{r}
qobj <- qvalue::qvalue(kern2[,1])
FDR[1,6] <- length(qobj$qvalues[qobj$qvalues < 0.1])
qobj <- qvalue::qvalue(gshoot2[,1])
FDR[2,6] <- length(qobj$qvalues[qobj$qvalues < 0.1])
qobj <- qvalue::qvalue(groot2[,1])
FDR[3,6] <- length(qobj$qvalues[qobj$qvalues < 0.1])
qobj <- qvalue::qvalue(base2[,1])
FDR[4,6] <- length(qobj$qvalues[qobj$qvalues < 0.1])
qobj <- qvalue::qvalue(tip2[,1])
FDR[5,6] <- length(qobj$qvalues2[qobj$qvalues < 0.1])
qobj <- qvalue::qvalue(lmad82[,1])
FDR[6,6] <- length(qobj$qvalues2[qobj$qvalues < 0.1])
qobj <- qvalue::qvalue(lman82[,1])
FDR[7,6] <- length(qobj$qvalues2[qobj$qvalues < 0.1])
qobj <- qvalue::qvalue(lmad262[,1])
FDR[8,6] <- length(qobj$qvalues[qobj$qvalues < 0.1])
qobj <- qvalue::qvalue(lman262[,1])
FDR[9,6] <- length(qobj$qvalues[qobj$qvalues < 0.1])

```

Reshape Data for Plotting 
```{r}
FDR$Tissue <- c("Kern", "GShoot", "GRoot", "L3Base", "L3Tip", "LMAD8", "LMAN8", "LMAD26", "LMAN26")
dat <- melt(FDR, id.vars = "Tissue")
dat[dat == 0] <- NA

```

Plot figure 2A
```{r}
pl <- ggplot(data=dat,aes(x=variable,y=Tissue)) + 
  geom_tile(aes(fill=value),color='black') + scale_fill_gradient(low = 'lightyellow', high = "#CC79A7", guide = FALSE, na.value = "white") + theme_bw() + xlab("\n") + ylab("Tissue") + 
  theme(axis.ticks=element_blank(),panel.border=element_blank(),panel.grid.major = element_blank(), axis.text.y = element_text(size=10,angle=0), axis.title.y = element_text(size=16),axis.title.x  = element_text(size=16),axis.text.x = element_text(angle = 0, hjust = 0.5,size=14)) + geom_text(aes(label=value),colour="grey15",size=3.5)

pl
```

Plot figure 2B 

First read in population and expression information
```{r}
pop <- read.csv("../data/FlintGarciaTableS1_fixednew.csv")
kern_all <- read.table("../data/Mean_centered_expression/Kern.txt")
rows <- row.names(kern_all)
kern_all$lines <- rows
pop_dat <- merge(pop, kern_all, by.x = "Inbred", by.y = "lines" )
```

Get Eigen values and vectors 
```{r}
## Kinship Matrix for all 207 Kern lines
myF <- read.table('../data/Kinship_matrices/F_Kern.txt')

## Get Eigen Values and Vectors 
myE <- eigen(myF)
E_vectors <- myE$vectors
E_values <- myE$values
```

Expression Plot
```{r}
pop_dat$sel <- kern_all[,1706]
pop_dat$one <- E_vectors[,1]


lambda <- E_values[1]
var_C11_20 <- 190.0949 #Expected denominator calculated from script 
k <-  1.96 * sqrt(var_C11_20 * lambda) 

k_plus <- k 
k_minus <- -k 

lR <- lm(pop_dat$sel ~ pop_dat$one) 
coeff <- lR$coefficients[[2]]

col <- c('#E69F00', '#56B4E9', "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
pl1 <- ggplot(data=pop_dat, aes(x = one, y= sel , color=Subpopulation)) + scale_colour_manual(values = col, labels=c("mixed", "non-stiff stalk", "popcorn", "stiff stalk", "sweet", "tropical")) + xlab("PC 1") +  ylab("Expression") + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.title.y = element_text(size=12), axis.title.x  = element_text(size=12), legend.position = "right", legend.title = element_text(size = 12), legend.text = element_text(size = 10), plot.title = element_text(hjust = 0.5)) + geom_point(size = 2.5) + geom_abline(slope = k_plus, linetype = 2) + geom_abline(slope = coeff, size = 1.5)+  geom_abline(slope = k_minus, linetype = 2) + ggtitle("GRMZM2G038195")

pl1
```

Plot Figure 2C 
```{r}
pop_dat$two <- E_vectors[,2]
pop_dat$sel <- kern_all[,1867]

lambda <- E_values[2]
# Expected denominator calculated from script 
k <-  1.96 * sqrt(20.96418 * lambda) 

k_plus <- k 
k_minus <- -k 

lR <- lm(pop_dat$sel ~ pop_dat$two) 
coeff <- lR$coefficients[[2]]

pl2 <- ggplot(data=pop_dat, aes(x = two, y= sel , color=Subpopulation)) + scale_colour_manual(values = col, labels=c("mixed", "non-stiff stalk", "popcorn", "stiff stalk", "sweet", "tropical")) + xlab("PC 2") +  ylab("Expression") + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.title.y = element_text(size=12), axis.title.x  = element_text(size=12), legend.position = "right", legend.title = element_text(size = 12), legend.text = element_text(size = 10),plot.title = element_text(hjust = 0.5)) + geom_point(size = 2.5) + geom_abline(slope = k_plus, linetype = 2) + geom_abline(slope = coeff, size = 1.5)+  geom_abline(slope = k_minus, linetype = 2) + ggtitle('GRMZM2G042055')

pl2
```

Final Figure  
```{r, warning=FALSE}
final <- ggarrange(pl,                                                 
          ggarrange(pl1, pl2, ncol = 2, labels = c("B", "C"),common.legend = T, legend = "bottom"), 
          nrow = 2, 
          labels = "A"                                       
          )
final
```

