---
title: "Detecting selection on expression of individual genes"
author: "Emily Josephs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(reshape2)
library(ggpubr)
library(quaint)
library(tidyr)
```


## Intro 

Here we identify genes whose expression is under divergent selection along the first 5 principal components in the dataset using the R package QUAINT. We have data for 7 different tissue types, the number of lines and genes per tissue is recorded in table S1. The mean centered expression values for the genes for each tissue type are stored in "data/Mean_centered_expression/*tissuename*.txt". The kinship matrices, generated from 78,324 randomly chosen SNPs for each tissue type, are stored in "../data/Kinship_matrices/F_*tissuename*".      

## Code 

This function takes in a tissue name as an argument and returns the results of calcQpc() for each gene. calcQpc() takes a vector of traits ($Z_1, Z_2, ...Z_m$), the matrix of eigenvectors ($\vec{U_1}, \vec{U_2}, ...\vec{U_m}$) and vector of eigenvalues ($\lambda_1, \lambda_2, ... \lambda_m$) generated form the eigen decomposition of the kinship matrix, the range of PCs used to estimate $V_a$ and the range of PCs you want to test for selection. Here our trait values are the mean centered expression values and we use the last half of PCs to estimate $V_a$ and test for selection along the first 5 PCs. calcQpc() outputs a list that includes the $c_m$ values (see equation 2) for the 5 PCs we are testing for selection along with the p-value from the F-test (see equation 3). The testAllGenes function runs calcQpc() on all genes and outputs the above information for every gene.  

```{r}
############ function for testing all genes
testAllGenes <- function(myTissue){
# Read in mean-centered expression values
df1 <- read.table(paste("../data/Mean_centered_expression/",myTissue,".txt",sep=""))
geneNames = names(df1)
# Read in tissue specific kinship matrix 
myF <- read.table(paste('../data/Kinship_matrices/F_',myTissue,'.txt',sep=""))

## Get Eigen Values and Vectors 
myE <- eigen(myF)
E_vectors <- myE$vectors
E_values <- myE$values

## Testing for selection on first 5 PCs
myM = 1:5

## Using the last 1/2 of PCs to estimate Va
myL = 6:dim(myF)[1]

## test for selection on each gene
allGeneOutput <- sapply(1:ncol(df1), function(x){
myQpc = calcQpc(myZ = df1[,x], myU = E_vectors, myLambdas = E_values, myL = myL, myM = myM)
return(c(geneNames[x],myQpc))
})

return(allGeneOutput)
}
```

Now we will run the test for selection on all of our tissue types. 
```{r}
###run on all genes
alltissues = c('GRoot',"Kern","LMAD26","LMAN26", "L3Tip","GShoot","L3Base")
alltissueresults = lapply(alltissues, testAllGenes)
names(alltissueresults) <- alltissues
```

Now lets look at the number of gene with FDR < 0.01 for each tissue
```{r}
##look at sig results
sigresults = lapply(1:length(alltissues), function(i){
# extract the pvalues
pvals = matrix(unlist(alltissueresults[[i]][5,]), ncol=5, byrow=TRUE) #each row corresponds to a gene, columns are to PCs

##look at pvalue distributions
# par(mfrow=c(3,2), mar=c(4,4,1,1))
# sapply(1:5, function(x){
# hist(pvals[,x], border="white", col = "darkgray", main="", breaks = 20, xlab = paste('PC ',x,' ',alltissues[i],sep=""))
# })

## test fpr significance with the fdr
pfdr = data.frame(apply(pvals,2, function(x){p.adjust(x, method='fdr')}))

## how many are significant?
numsig <- apply(pfdr, 2, function(x){sum(x<0.1)})
numsig
})

###make a big image of how many sig results we have
sigTable = as.data.frame(matrix(unlist(sigresults), ncol=5, byrow=T))
names(sigTable) = c('PC1','PC2','PC3','PC4','PC5')
sigTable$nicename = c('Germinating root', 'Kernel','Adult Leaf Day', 'Adult Leaf Night', '3rd leaf tip', 'Germinating shoot','3rd leaf base')
sigTable

sigLong = tidyr::gather(sigTable, 'variable','value', -nicename)
sigLong[sigLong == 0] <- NA

###save output for other analyses
save(alltissueresults, sigTable, sigLong, file = "../data/quaint-results.rda")
```

We can also check how many of our significant genes are unique. Here 60 of 66 significant genes are unqiue to their tissue/PC combination.  
```{r}
#how many unique genes???
siggenes = unlist(lapply(1:length(alltissues), function(i){
# extract the pvalues
pvals = matrix(unlist(alltissueresults[[i]][5,]), ncol=5, byrow=TRUE) #each row corresponds to a gene, columns are to PCs
pfdr = data.frame(apply(pvals,2, function(x){p.adjust(x, method='fdr')})) ## test fpr significance with the fdr
pfdr$numsig <- apply(pfdr, 1, function(x){sum(x<0.1)})
pfdr$genes = unlist(alltissueresults[[i]][1,])
return(dplyr::filter(pfdr, numsig>0)$genes)
}))
length(unique(siggenes)) ##number of unique genes
```

## Figures  

Now that we have a table with the number of significant genes per tissue. let's make a heat map of them with ggplot. This is Figure 1A in the paper.  
```{r}
pl <- ggplot(data=sigLong,aes(x=variable,y=nicename)) + 
  geom_tile(aes(fill=value),color='black') + scale_fill_gradient(low = 'lightyellow', high = "#CC79A7", guide = FALSE, na.value = "white") + theme_bw() + xlab("\n") + ylab("Tissue") + 
  theme(axis.ticks=element_blank(),panel.border=element_blank(),panel.grid.major = element_blank(), axis.text.y = element_text(size=10,angle=0), axis.title.y = element_text(size=16),axis.title.x  = element_text(size=16),axis.text.x = element_text(angle = 0, hjust = 0.5,size=14)) + geom_text(aes(label=value),colour="grey15",size=5.5)
pl
```

Make a table of the sample sizes - Table S1. 
```{r}
mySampleTable = sapply(alltissues, function(myTissue){
df1 <- read.table(paste("../data/Mean_centered_expression/",myTissue,".txt",sep=""))
numGenes = dim(df1)[2]
numInds = dim(df1)[1]
return(c(numGenes,numInds))
})
rownames(mySampleTable) = c('genes','individuals')

mySampleTable
```

Here we have a function that allow's us to plot expression vs PC
```{r}
### function for making plots
makeGenePlot <- function(tissue, geneIndex, pc){

## read in population data
pop <- read.csv("../data/FlintGarciaTableS1_fixednew.csv", stringsAsFactors = F)

## read in expression data
exp <- read.table(paste("../data/Mean_centered_expression/",tissue,".txt", sep=""), stringsAsFactors = F)
expgene = data.frame(Inbred = row.names(exp), geneexp = exp[,geneIndex], stringsAsFactors = F)

## merge data together
pop_dat <- dplyr::inner_join(pop, expgene, by= "Inbred")

#get the eigenvalues
myF <- read.table(paste('../data/Kinship_matrices/F_',tissue,'.txt', sep=""))
myE = eigen(myF)
myPC = data.frame(pc = myE$vectors[,pc], stringsAsFactors = F)
pop_dat <- dplyr::bind_cols(pop_dat, myPC)

lambda <- myE$values[pc]

##calculate the CIs
generesults = alltissueresults[[tissue]][,geneIndex]
myVaest = var0(generesults$cml)
myCI = 1.96*sqrt(myVaest*lambda)

##gene name
geneName = names(exp)[geneIndex]

lR <- lm(pop_dat$geneexp ~ pop_dat$pc) 
coeff <- lR$coefficients[[2]]

if(tissue=="LMAD26"){mylab = c("mixed", "non-stiff stalk", "popcorn", "stiff stalk", "tropical")} else{mylab = c("mixed", "non-stiff stalk", "popcorn", "stiff stalk", "sweet", "tropical")} ##no sweets in LMAD26

col <- c('#E69F00', '#56B4E9', "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
pl1 <- ggplot(data=pop_dat, aes(x = pc, y= geneexp , color=Subpopulation)) + scale_colour_manual(values = col, labels=mylab) + xlab(paste("PC",pc)) +  ylab("Expression") + theme_bw() + theme(panel.grid.major = element_blank(), text = element_text(size=15), panel.grid.minor = element_blank(), axis.title.y = element_text(size=18), axis.title.x  = element_text(size=18), legend.position = "right", legend.title = element_text(size = 12), legend.text = element_text(size = 20), plot.title = element_text(hjust = 0.5)) + geom_point(size = 3.5) + geom_abline(slope = myCI, linetype = 2, col = '#56B4E9', size=1.2) + geom_abline(slope = coeff, size = 1.5)+  geom_abline(slope = -myCI, linetype = 2, col = '#56B4E9', size=1.2)

return(pl1)
}

```

Here we will plot Figure 1B and 1C. The code bellow can be modified to make expression plots for different gene and different tissues by changing the tissue name in the first line and giving the index of the gene you want to plot. The index can be found by finding the column number 
```{r}
myTissue = 'LMAD26'
df1 <- read.table(paste("../data/Mean_centered_expression/",myTissue,".txt",sep=""))
geneNames = names(df1)  

## Pick name of gene to plot and plot vs PC 
gene_index <- which(geneNames == "GRMZM2G152686")
pc1_plot = makeGenePlot(tissue = myTissue,geneIndex = gene_index, pc = 1)
pc1_plot

## Pick name of gene to plot and plot vs PC 
gene_index <- which(geneNames == "GRMZM2G069762")
pc5_plot = makeGenePlot(tissue = myTissue,geneIndex = gene_index, pc = 5)
pc5_plot
```

Make Final Figure  
```{r}
final <- ggarrange(pl,                                                 
          ggarrange(pc1_plot, pc5_plot, ncol = 2, labels = c("B", "C"),common.legend = T, legend = "bottom"), 
          nrow = 2, 
          labels = "A"                                       
          )
final
```




