---
title: "Population Structure"
author: "Jennifer Blanc"
date: "2/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Intro   

Here is the code to create figure 1. 

## Code 

First we need to read in the file containing the 302 Maize lines in the data set and the mean centered expression values for the 207 lines that had expression data for the Kernel tissue. In this matrix each row corresponds to a line and each column is a gene. We then merge the tables by line name. 
```{r}
pop <- read.csv("data/FlintGarciaTableS1_fixednew.csv")
kern_all <- read.table("data/Mean_centered_expression/Kern.txt")
rows <- row.names(kern_all)
kern_all$lines <- rows
pop_dat <- merge(pop, kern_all, by.x = "Inbred", by.y = "lines" )
```

Next we read in the Kinship matrix for the 207 kernel lines and get the Eigen vectors and values. 
```{r}
## Kinship Matrix for all 207 Kern lines
myF <- read.table('data/Kinship_matrices/F_Kern.txt')

## Get Eigen Values and Vectors 
myE <- eigen(myF)
E_vectors <- myE$vectors
E_values <- myE$values
```


Finally we add the first and second eigen vectors to the dataframe that has the population and expression information.
```{r}
pop_dat$one <- E_vectors[,1]
pop_dat$two <- E_vectors[,2]
```

## Figure  
```{r}
col <- c('#E69F00', '#56B4E9', "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
pl <- ggplot(data=pop_dat, aes(x = one, y= two , color=Subpopulation)) + scale_colour_manual(values = col, labels=c("mixed", "non-stiff stalk", "popcorn", "stiff stalk", "sweet", "tropical")) + xlab("PC 1") +  ylab("PC 2") + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.title.y = element_text(size=17), axis.title.x  = element_text(size=17), legend.position = "bottom", legend.title = element_text(size = 12), legend.text = element_text(size = 10), plot.title = element_text(), axis.text.y = element_text(size=10), axis.text.x = element_text(size=10)) + geom_point(size = 1.5) 
pl
```

