---
title: "Selection on Expression of Cold Response Genes"
author: "Jennifer Blanc"
date: "2/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(knitr)
library(reshape2)
```

## Intro  

The goal of this analysis is to determine if genes that have been shown to be differentially expressed in response to cold treatment are enriched for genes whose expression is under selection. 


## Code 

First, read in the Avila et al. differential expression data and get the names of the genes used in their analysis. 
```{r}
cold <- read.csv("../data/Cold.csv", header = F, stringsAsFactors = F)
cold <-cold[,2:19] 
colnames(cold) <- c("V3_Gene_Name","V4_Gene_Name","Arabidopsis_Ortholog","Rice_Ortholog","Cold_1D_CG60_Means","Cold_1D_CG102_Means","Control_1D_CG60_Means","Control_1D_CG102_Means","FDR_1D_Genotype","FDR_1D_Treatment","FDR_\
1D_Interaction","Cold_4D_CG60_Means","Cold_4D_CG102_Means","Control_4D_CG60_Means","Control_4D_CG102_Means","FDR_4D_Genotype","FDR_4D_Treatment","FDR_4D_Interaction")

cold_names <- cold$V3_Gene_Name
```

We need to pick only the genes for which we have expression data for. To do this we will load the names for the genes we have for each tissue and take the overlap with the Avila data.  
```{r}

alltissues = c('GRoot',"Kern","LMAD26","LMAN26", "L3Tip","GShoot","L3Base")

load('../data/quaint-results.rda')

```

The next step is to determine if differentially expressed genes are more likely to be under selection in every PC/Tissue combination. To do this we will use a chi-squared test.  

```{r, warning=FALSE, message=FALSE}
coldsum <- cold %>% select(V3_Gene_Name, V4_Gene_Name, FDR_1D_Treatment, FDR_4D_Treatment) ## pull out important columns

coldsum$sig = apply(coldsum[,3:4], 1, min, na.rm=T)
coldsigGenes = dplyr::filter(coldsum, sig < 0.1)$V3_Gene_Name
length(coldsigGenes)
coldnotsigGenes = dplyr::filter(coldsum, sig > 0.1)$V3_Gene_Name
length(coldnotsigGenes)


##function for running the chisq test on a specific pc/tissue combination
run_chi_sq <- function(mytissue, myPC,sigGenes, notsigGenes, myp = 0.05) {
  quaint_results = alltissueresults[[mytissue]] ##pull out tissue specific genes
  quaint_results_pc = data.frame(mygene = unlist(quaint_results[1,]), pval = unlist(lapply(quaint_results[5,], function(x){x[myPC]})), stringsAsFactors = F)  ## just need a table for the specific PC's pvalue
  sigOver = dplyr::filter(quaint_results_pc, mygene %in% sigGenes)
  notsigOver = dplyr::filter(quaint_results_pc, mygene %in% notsigGenes)
  ##make the chisq table
  bothsig = nrow(dplyr::filter(sigOver, pval < myp))
  diffexponly = nrow(dplyr::filter(sigOver, pval >= myp))
  selonly = nrow(dplyr::filter(notsigOver, pval < myp))
  nothingsig = nrow(dplyr::filter(notsigOver, pval > myp))
  chitbl = matrix(c(bothsig, selonly, diffexponly, nothingsig), nrow=2) ##make the chisq table
  chitest = chisq.test(chitbl, simulate.p.value = TRUE) ##run the test
  return(chitest$p.value)
}

##run on all PCs and all tissues for the cold genes
myresults = sapply(alltissues, function(mytissue){
  sapply(1:5,function(myPC){
    run_chi_sq(mytissue, myPC, coldsigGenes, coldnotsigGenes, 0.05)
  })
})


##we really just care about combos with significant evidence of selection in at least one gene

newresults = myresults
newresults[t(sigTable[,1:5]) ==0] <- NA
newresults*15

### note that we don't know what direction this goes in.

```

## Drought response genes
We use drought response genes from Forestan et al. (https://onlinelibrary-wiley-com.proxy2.cl.msu.edu/doi/full/10.1111/pce.13660)
Only looking at NST0 vs WST0 because that shows comparison of expression changes in drought and control.
NSTO is control, WSTO is stressed


```{r}
droughtgenes = read.csv('../data/pce13660-supp-0004-supplementalfile3_cuffdiff.csv', stringsAsFactors = F)
droughtgenes$v4_gene_model = sapply(droughtgenes$gene_id, function(x){strsplit(x, split=':')[[1]][2]}) #make a column for gene names

#convert gene names from v4 to v3
genemodel = read.csv('../data/gene_model_xref_v3.csv', stringsAsFactors = F)[,1:2]
dgmerge = dplyr::inner_join(droughtgenes, genemodel, by = "v4_gene_model")
dim(dgmerge)

droughtsig = dplyr::filter(dgmerge, q_value < 0.1)$v3_gene_model ##filter out sig genes
length(droughtsig)
droughtnot = dplyr::filter(dgmerge, q_value > 0.1)$v3_gene_model ##filter out sig genes
length(droughtnot)
```

```{r, eval=F, echo=F}
###not controlling for expression

dresults = run_chi_sq(mytissue = 'LMAD26', myPC=5, sigGenes = droughtsig, notsigGenes = droughtnot, myp = 0.05)

dresults = sapply(alltissues, function(mytissue){
  sapply(1:5,function(myPC){
    run_chi_sq(mytissue, myPC, droughtsig, droughtnot, 0.05)
  })
})

dresults[t(sigTable[,1:5]) ==0] <- NA #remove thiings with no significant signal of selection


dresults
dresults*15

## robust to pvalues?
#dresults = sapply(alltissues, function(mytissue){
#  sapply(1:5,function(myPC){
#    run_chi_sq(mytissue, myPC, droughtsig, droughtnot, 0.01)
#  })
#})
#dresults[t(sigTable[,1:5]) ==0] <- NA #remove thiings with no significant signal of selection
#dresults


##compare up and down regulated genes
drought_up = dplyr::filter(dgmerge, value_1 < value_2 & q_value < 0.1)$v3_gene_model
drought_down = dplyr::filter(dgmerge, value_1 > value_2 & q_value < 0.1)$v3_gene_model
length(drought_up)
length(drought_down)

##test for enrichment
dresults_up = sapply(alltissues, function(mytissue){
  sapply(1:5,function(myPC){
    run_chi_sq(mytissue, myPC, drought_up, droughtnot, 0.05)
  })
})

dresults_up[t(sigTable[,1:5]) ==0] <- NA #remove thiings with no significant signal of selection
dresults_up


dresults_down = sapply(alltissues, function(mytissue){
  sapply(1:5,function(myPC){
    run_chi_sq(mytissue, myPC, drought_down, droughtnot, 0.05)
  })
})

dresults_down[t(sigTable[,1:5]) ==0] <- NA #remove thiings with no significant signal of selection
dresults_down
```

```{r}
##comparing overall expression level in the Kremling data between categories (look at LMAD26 for now)

mytissue = 'LMAD26'

exp = read.table(paste('../data/Raw_expression/',mytissue,'.txt', sep=''), stringsAsFactors = F)
exp_means = data.frame(gene = colnames(exp), expr = colMeans(exp), stringsAsFactors = F, row.names=NULL)

quaint_results = alltissueresults[[mytissue]]
quaint_ps = data.frame(gene = unlist(quaint_results[1,]),matrix(unlist(quaint_results[5,]), ncol=5, byrow=T), stringsAsFactors = F)
names(quaint_ps) =c('gene','PC1','PC2','PC3','PC4','PC5')
quaint_mer = dplyr::inner_join(exp_means, quaint_ps, by = 'gene') #merge with expression data

mydflong = tidyr::gather(quaint_mer, 'PC','pval', PC1:PC5)

##compare expression levels
mysig = dplyr::filter(mydflong, pval < 0.05)
mynot = dplyr::filter(mydflong, pval >= 0.05)

## compare expression levels of diff expressed genes
droughtmer = dplyr::inner_join(dgmerge, exp_means, by=c('v3_gene_model' = 'gene'))
dsig = dplyr::filter(droughtmer, q_value < 0.1)
dnot = dplyr::filter(droughtmer, q_value >=0.1)

par(mar=c(8,5,2,2))
plot(c(1,2,4,5), c(median(mysig$expr), median(mynot$expr), median(dsig$expr), median(dnot$expr)),
     bty="n", ylim=c(0,180), xaxt="n", yaxt="n",ylab = "median expression in leaves", xlab="", xlim = c(0,6))
axis(1, at=c(1,2,4,5), lab = c('selection','not selection','drought response','no response'), las=2)
axis(2,las=2)
###
```

```{r}
## pick the top  expressed genes in dnot 
dnot_high = dnot[order(-dnot$expr),][1:3500,]

#down regulated genes
dsig_down = dplyr::filter(dsig, value_1 > value_2)
dsig_up = dplyr::filter(dsig, value_2 > value_1)
nrow(dsig_down)
nrow(dsig_up)

par(mar=c(8,5,2,2))
plot(c(1,2,4,5,6,7), c(median(mysig$expr), median(mynot$expr), median(dsig$expr), median(dsig_down$expr), median(dnot$expr), median(dnot_high$expr)),
     bty="n", ylim=c(0,180), xaxt="n", yaxt="n",ylab = "median expression in leaves", xlab="", 
     xlim = c(0,8)) 
axis(1, at=c(1,2,4,5,6,7), 
    lab = c('selection','not selection','drought response','drought down','no response','no, high'), las=2)
axis(2,las=2)

##do the test with only things that have leaf expression in Kremling et al
dresults_leaf = sapply(alltissues, function(mytissue){
  sapply(1:5,function(myPC){
    run_chi_sq(mytissue, myPC, dsig$v3_gene_model, dnot$v3_gene_model, 0.05)
  })
})
dresults_leaf[t(sigTable[,1:5]) ==0] <- NA #remove thiings with no significant signal of selection
dresults_leaf*15

###now test for enrichment compared to the high expression sample
dresults_high = sapply(alltissues, function(mytissue){
  sapply(1:5,function(myPC){
    run_chi_sq(mytissue, myPC, dsig$v3_gene_model, dnot_high$v3_gene_model, 0.05)
  })
})
dresults_high[t(sigTable[,1:5]) ==0] <- NA #remove thiings with no significant signal of selection
dresults_high*15

##just down reg responses
dresults_high_down = sapply(alltissues, function(mytissue){
  sapply(1:5,function(myPC){
    run_chi_sq(mytissue, myPC, dsig_down$v3_gene_model, dnot_high$v3_gene_model, 0.05)
  })
})
dresults_high_down[t(sigTable[,1:5]) ==0] <- NA #remove thiings with no significant signal of selection
dresults_high_down*15

##up responses
dresults_high_up = sapply(alltissues, function(mytissue){
  sapply(1:5,function(myPC){
    run_chi_sq(mytissue, myPC, dsig_up$v3_gene_model, dnot_high$v3_gene_model, 0.05)
  })
})
dresults_high_up[t(sigTable[,1:5]) ==0] <- NA #remove thiings with no significant signal of selection
dresults_high_up*15
```




## Figure  
  
```{r, eval=T}
make_enrichment_fig <- function(mytissue, myPC, myp, sigGenes, notsigGenes){
quaint_results = alltissueresults[[mytissue]] ##pull out tissue specific genes
quaint_results_pc = data.frame(mygene = unlist(quaint_results[1,]), pval = unlist(lapply(quaint_results[5,], function(x){x[myPC]})), stringsAsFactors = F)  ## just need a table for the specific PC's pvalue
sigOver = dplyr::filter(quaint_results_pc, mygene %in% sigGenes)
notsigOver = dplyr::filter(quaint_results_pc, mygene %in% notsigGenes)
##make the chisq table
bothsig = nrow(dplyr::filter(sigOver, pval < myp))
diffexponly = nrow(dplyr::filter(sigOver, pval >= myp))
selonly = nrow(dplyr::filter(notsigOver, pval < myp))
nothingsig = nrow(dplyr::filter(notsigOver, pval > myp))
  
propDiffSel = bothsig/(bothsig+diffexponly)
propNotDiffSel = selonly/(selonly+nothingsig)

return(c(propDiffSel, propNotDiffSel))
}

test = make_enrichment_fig('LMAD26', 5, 0.05, coldsigGenes, coldnotsigGenes)
bp = barplot(test, ylim=c(0,0.1), col = c("#0072B2","#CC79A7"), border="white", yaxt = "n", ylab = "Proportion under selection", cex.lab=1.5)
axis(2, las=2)
axis(1, at = bp, lab = c('Cold Response','Other'), cex.axis=1.5)


#### drought figure
test = make_enrichment_fig('LMAD26', 5, 0.05, dsig_down$v3_gene_model, dnot_high$v3_gene_model)
bp = barplot(test, ylim=c(0,0.15), col = c("#0072B2","#CC79A7"), border="white", yaxt = "n", ylab = "Proportion under selection", cex.lab=1.5)
axis(2, las=2)
axis(1, at = bp, lab = c('Drought Response','Other'), cex.axis=1.5)


test = make_enrichment_fig('LMAN26', 2, 0.05, dsig_down$v3_gene_model, dnot_high$v3_gene_model)
bp = barplot(test, ylim=c(0,0.1), col = c("#0072B2","#CC79A7"), border="white", yaxt = "n", ylab = "Proportion under selection", cex.lab=1.5)
axis(2, las=2)
axis(1, at = bp, lab = c('Drought Response','Other'), cex.axis=1.5)



get_enrichment <- function(mytissue, myPC, myp, sigGenes, notsigGenes){
quaint_results = alltissueresults[[mytissue]] ##pull out tissue specific genes
quaint_results_pc = data.frame(mygene = unlist(quaint_results[1,]), pval = unlist(lapply(quaint_results[5,], function(x){x[myPC]})), stringsAsFactors = F)  ## just need a table for the specific PC's pvalue
sigOver = dplyr::filter(quaint_results_pc, mygene %in% sigGenes)
notsigOver = dplyr::filter(quaint_results_pc, mygene %in% notsigGenes)
##make the chisq table
bothsig = nrow(dplyr::filter(sigOver, pval < myp))
diffexponly = nrow(dplyr::filter(sigOver, pval >= myp))
selonly = nrow(dplyr::filter(notsigOver, pval < myp))
nothingsig = nrow(dplyr::filter(notsigOver, pval > myp))
return(c(bothsig, diffexponly))
}



sigcounts = sapply(alltissues, function(mytissue){
  sapply(1:5,function(myPC){
    get_enrichment(mytissue, myPC, 0.05,dsig_down$v3_gene_model, dnot_high$v3_gene_model)
  })
})



```

